Type Visualization for R

### 2020-06-20 paper/main.tex
paper: Finish up

%% Paper title.

@@

%% This is how authors are specified in the conference style

%% Author and Affiliation (single author).
%%\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com}}
%%\affiliation{\scriptsize Allied Widgets Research}

%% Author and Affiliation (multiple authors with single affiliations).
%%\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com} %
%%\and Ed Grimley\thanks{e-mail:ed.grimley@aol.com} %
%%\and Martha Stewart\thanks{e-mail:martha.stewart@marthastewart.com}}
%%\affiliation{\scriptsize Martha Stewart Enterprises \\ Microsoft Research}

%% Author and Affiliation (multiple authors with multiple affiliations)
\author{Roy G. Biv\thanks{roy.g.biv@aol.com}\\ %
        \scriptsize Starbucks Research %
\and Ed Grimley\thanks{ed.grimley@aol.com}\\ %
     \scriptsize Grimley Widgets, Inc. %
\and Martha Stewart\thanks{martha.stewart@marthastewart.com}\\ %
     \parbox{1.4in}{\scriptsize \centering Martha Stewart Enterprises \\ Microsoft Research}}

%% A teaser figure can be included as follows, but is not recommended since
%% the space is now taken up by a full width abstract.
%\teaser{
%  \includegraphics[width=1.5in]{sample.eps}
%  \caption{Lookit! Lookit!}
%}

%% Abstract section.
\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built {\sc TypeVis} to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  {\sc TypeVis} uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.

  \todo{Fill in keywords below.}
} % end of abstract

@@

\section{Background and Related Work}

@@

\todo{Prior work}

@@

@@

Prominently featured as the center visualization,
a Sankey diagram encodes type signatures of the selected
functions as flows.
Flows begin at nodes labeling a function,
curve across each argument type in order,
and terminate at the return type.
Widths are proportional to how many times a function
is called with that signature during analysis.
Flow segments are filled with a hue-varying color scale,
based on the type where the segment ends.
Beneath each node label is an approximate count.
If the node represents a function, then the
number is how many times the function was invoked.
If the node represents a type, then the
number is how often a value of that type
at that argument position was recorded during analysis.

@@

Interactivity significantly augments the
usability of the Sankey visualization.
Hovering over a flow will highlight the
entire path of the flow
and will fade other functions
into the background.
When displaying many flows at the same time,
highlighting becomes especially critical
for user comprehension.
Additional quantitative information
about the flow is also supplied
while hovering.
Clicking a flow will focus on that function
only, filtering away all others.

@@

% other dynamic languages

% other featuers not just types

% general philosophy not just about types
% vis has role to play in language design

\todo{Write.}

\acknowledgments{
  The authors wish to thank A, B, and C. This work was supported in part by
  a grant from XYZ.
}

>>>>

@@

\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com} %
\and Ed Grimley\thanks{e-mail:ed.grimley@aol.com} %
\and Martha Stewart\thanks{e-mail:martha.stewart@marthastewart.com}}
\affiliation{\scriptsize Northeastern University}

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built {\sc TypeVis} to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  {\sc TypeVis} uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.

  \todo{Fill in keywords below.}
}

@@

\section{Background and Prior Work}

@@

We are not aware of any previous work that applies
visualization techniques for the purposes of programming language design.
However, there are existing publications on understanding
software systems through visualization.
Bohnet visualize massive program execution traces
to assist developers in code comprehension of large software systems.
Their tool applies pruning and summarization to reduce trace size.
The visualization focuses on call topology,
describing the structure of function invocations,
rather than types.
Telea also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
Our visualization can be thought of as a hybrid of the two,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza adorn nodes and edges with extra information,
allowing the user gain further understanding of the system
under examination.
This includes indicating methods that come from the same class---a
basic form of type information.

@@

Treemaps, introduced by Shneiderman,
are often used for displaying hierarchical data.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the dataset was more deeply nested,
{\sc TypeVis} can be configured to
handle that.

@@

Prominently featured as the center visualization,
a parallel sets diagram encodes type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara,
are ideally suited for type signatures.
Each data point is can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n)$
corresponding to the function name,
argument types,
and return type.
Every component is categorical,
and we are interested in visualizing frequencies of these points.
This is the exact use-case for parallel sets.

Flows begin at nodes labeling a function,
curve across each argument type in order,
and terminate at the return type.
Widths are proportional to how many times a function
is called with that signature during analysis.
Flow segments are filled with a hue-varying color scale,
based on the type where the segment ends.
Beneath each node label is an approximate count.
If the node represents a function, then the
number is how many times the function was invoked.
If the node represents a type, then the
number is how often a value of that type
at that argument position was recorded during analysis.

@@

Interactivity significantly augments the
usability of the parallel sets visualization.
Hovering over a flow will highlight the
entire path of the flow
and will fade other functions
into the background.
When displaying many flows at the same time,
highlighting becomes especially critical
for user comprehension.
Additional quantitative information
about the flow is also supplied
while hovering.
Clicking a flow will focus on that function
only, filtering away all others.

@@

We have presented {\sc TypeVis}, an interactive visualization
of run-time type information for R.
It supports real-time exploration of a dataset of
execution traces from a large corpus of R packages.
To narrow the data under consideration,
{\sc TypeVis} provides facilities for filtering
and searching.
Type signatures are represented as flows travesing across
each argument type and return type in sequence.
Flow widths are proportional to how often
that type signature was seen during the dynamic analysis.
Interactions provide additional details and
assist users in comprehension and navigation.

Our tool is aimed at elucidating how R programmers
use the language in practice.
Specifically, the data and visualization will be used to
inform the design of a gradual type system for R.
However, {\sc TypeVis} is not tied to the R ecosystem
and is suitable to use for any dynamically typed language
with a similar dataset.
Additionally, the visualization could yield insights
useful for other purposes than gradual typing.

We hope that, in the future, programming language designers
will collect and use empirical data about how their languages
are used in practice.
Such data can be used to make informed decisions about
language design.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
{\sc TypeVis} is just the first step in this direction.

\todo{Finish these.}
\acknowledgments{
  The authors wish to thank the following people:
  students of CS 7250 for useful design feedback,
  contributors to all the open source projects that {\sc TypeVis} depends on.
  This work was supported in part by a Northeastern Graduate Fellowship.
}
